---
title: "COMBINE R Workshop Session 2"
author: Jia Qi Beh
output: html_document
date: "2023-02-10"
editor_options: 
  markdown: 
    wrap: sentence
---

# Basics of Data Manipulation: Part 1

In this session, we will look at how you can manipulate your data in R Studio.

The most commonly used data type is data frame, which are usually imported from CSV files.
Examples: .csv

## Install and load packages

For this session, we will be using the 'dplyr' package in tidyverse.

```{r}

install.packages("tidyverse")   # if you have not already done so 
library(tidyverse)   # load package 

```

You can check if the 'dplyr' was loaded into R from the 'Packages' panel on the right.

## Load dataset

We will use the "read.csv" function to load our data.

Hint: I prefer using the absolute or full path in the file name.
You can press TAB to navigate to your working directory.

```{r read.csv}

data <- read.csv("./s2_p1_data.csv")

data
View(data)  # opens a new window

```

We can extract information of the data frame with these functions:

```{r}

nrow(data)  # total number of rows
ncol(data)  # total number of columns 
dim(data)  # prints dimension - number of rows and columns
colnames(data)  # print all column names 
typeof(data$ID)  # check data type for each column
str(data)  # display data type of all columns 
summary(data$Age)  # prints summary of each data column

```

## Logical operators

-   `x == y` -- "equal to" (not the same as = )

-   `x != y` -- "not equal to"

-   `x < y` -- "less than"

-   `x > y` -- "greater than"

-   `x <= y` -- "less than or equal to"

-   `x >= y` -- "greater than or equal to"

-   `a & b` -- "and", TRUE only if both `a` and `b` are TRUE.

-   `a | b` -- "or", TRUE if either `a` or `b` or both are TRUE.

## Pipe operator

-   %\>% -- piping one function into the next - when combining two or more functions.

## mutate()

Now let's move on to the more exciting part of dplyr!

What if you want to create a new column which shows the difference between Average Age and Actual Age?

Hints:

1\.
We know the average age from summary( ) -- look for mean Age!

2\.
Use mutate() to create a new column.
(not row!)

```{r}

# look for differences between Mean Age and Age of individual
mutate(data, newcol = 26.23 - Age) 
# OR
data %>% mutate(newcol = 26.23 - Age)

# Save it to a new variable
data.1 <- data %>% mutate(newcol = 26.23 - Age)
data.1

data  # preserve the original data frame

```

Saving the new data to a new variable prevents the loss of the old dataset.
You can still call out the previous data using "data" or "View(data)".

## rename()

Next, you want to rename your new column.
But how?

1.  Use the rename() function.
2.  Be careful of the order -- NEW name = OLD name

```{r}

data.1 %>% rename(diffAge = newcol)

data.2 <- data.1 %>% rename(diffAge = newcol)
data.2

data.1

```

## relocate()

You wish to relocate the new column next to Age.

```{r}

data.2 
data.2 %>% relocate(diffAge, .after = Age)
# OR
data.2 %>% relocate(diffAge, .before = ReasonForAdmission)

data.2 <- data.2 %>% relocate(diffAge, .after = Age)  # replace original data frame
data.2

```

## group_by %\>% ungroup

What if you want to know the average Length of Stay by sex?

1.  Use group_by()
2.  After grouping, use summarise()

```{r}

data.2 %>% 
  group_by(Sex) %>% 
  summarise(n = n(),   # tells the total number of observations for each sex
            m = mean(LengthOfStay)) %>%   # calculates mean Age for each sex 
  ungroup 

```

Don't get mixed up with summarise() and summary()!

## filter()

What if you are only interested in the average Length of Stay among males?

1.  Use filter() to extract rows for Sex=Male
2.  Use summarise()

```{r}

data.2 %>% 
  filter(Sex == "Male")

data.2 %>% 
  filter(Sex == "Male") %>%    # be careful to use "" when specifying for object 
  summarise(m = mean(LengthOfStay))

```

How about looking at the average Length of Stay for males with Chronic illness?

Hint: Use the AND operator.

```{r}

data.2 %>% 
  filter(Sex == "Male" & ReasonForAdmission == "Chronic")

data.2 %>% 
  filter(Sex == "Male" & ReasonForAdmission == "Chronic") %>%  
  summarise(m = mean(LengthOfStay))

```

Taking it one step further - how would you find out the average Length of Stay for Males or Females with Chronic illness?

```{r}

data.2 %>%
  filter(Sex == "Male" | Sex == "Female", ReasonForAdmission == "Chronic") %>%
  summarise(m = mean(LengthOfStay))

```

Other options are available.
Feel free to play around with different logical operators.

What would you expect to get from the following examples?

```{r}

data.2 %>% filter(Sex != "Male")

data.2 %>% filter(Age > 30)   # Make sure Age is a number, not a Character

data.2 %>% filter(ReasonForAdmission %in% c("Chronic", "Accident"))  
# what does %in% mean?

data.2 %>% filter((Sex == "Female") | (Sex == "Male" & ReasonForAdmission == "Chronic"))

```

## select()

Previously, we use filter() to select one or more ROWS.

To select for one or more COLUMNS, we can use the 'select()' function.

```{r}

data.2 %>% select(ID, Sex, Age)

data.2 %>% select(1,2,3)
data.2 %>% select(1:3)

data.2 %>% select(-Sex)
data.2 %>% select(-c(Sex, Age))
data.2 %>% select(-(2:3))

```

This can be useful when you want to recreate a new data frame without certain columns.

## recode()

If you wish to rename the objects in your data, use recode().
NOTE: Again, be careful of the order -- "OLD object" = "NEW object"

```{r}

data.2 %>% mutate(Sex = recode(Sex, "Male" = "Man"))

data.2 %>% mutate(ReasonForAdmission = recode(ReasonForAdmission, 
                                              "Chronic" = "Cancer",
                                              "Emergency" = "ICU"))

```

Remember to save your new data frame into a variable after making changes!

## arrange()

This is useful when you want to re-order your data frame.

```{r}

data.2 %>% arrange(Age)

data.2 %>% arrange(desc(Age))

```

# Basics of Data Manipulation: Part 2

In Part 2, we will try out some of the commands that we have looked at previously, using a larger dataset.

For this, we are going to use the 'gapminder' dataset available from the R package gapminder.

To do this, you first have to install and load the package:

```{r}

install.packages("gapminder")
library(gapminder)

```

After loading the package, you can then call out the gapminder dataset.

NOTE: The gapminder dataset is already in a data frame, so you don't have to use 'read.csv'.

There are also other datasets available in Base R. Go to 'Packages' and look for the 'datasets' package.
After clicking on it, a 'Help' panel will pop up with the names of the dataset available.
For instance, the most commonly used dataset was the 'iris' dataset.

# Tasks:

## Warm-up

Can you tell how many rows and columns are there?
What are the column names?

Solution:

```{r}

gapminder  # view data frame

dim(gapminder)
colnames(gapminder)

```

## Question 1

You wish to know how many different countries were listed under each continent.
Also, you wanted to know the number of observations for each country.

Solution:

```{r}
#Type your answer here


```

## Question 2

You wish to compare the average life expectancy of the population in Europe before and after 2000.

Solution:

```{r}
#Type your answer here


```

## Question 3

You wish to know in which year did Australia and China record the highest gdp per capita.
Which country has the higher 'highest' gdp per capita?

Solution:

```{r}
#Type your answer here


```
