---
title: "COMBINE R Workshop 2022 - Vizualization"
author: "Mikhail Dias"
date: "19/02/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Plotting with ggplot2**

`ggplot2` is mini-language within R, a language for creating plots.

It implements ideas from a book called ["The Grammar of Graphics"](https://www.amazon.com/Grammar-Graphics-Statistics-Computing/dp/0387245448).

The syntax can be a little strange, but there are plenty of examples in the [online documentation](http://ggplot2.tidyverse.org/reference/).

`ggplot2` is part of the Tidyverse, so loading the `tidyverse` package will load `ggplot2`.

```{r message=F, warning=F}
#install.packages('tidyverse')
library(tidyverse)
```

Load the Gapminder dataset - make sure you include the (full) path to where they are saved:

```{r message=F, warning=F}
geo <- read_csv("r-intro-2-files/geo.csv")
gap <- read_csv("r-intro-2-files/gap-minder.csv")

gap_geo <- left_join(gap, geo, by="name")

head(gap_geo)

rm(geo, gap)
```

## Elements of a ggplot

Producing a plot with `ggplot2`, we must give three things:

1.  A data frame containing our data.
2.  How the columns of the data frame can be translated into positions, colors, sizes, and shapes of graphical elements ("aesthetics" = aka "aes").
3.  The actual graphical elements to display ("geometric objects").

Let's make our first ggplot.

```{r warning=F}
ggplot(data = gap_geo,
       aes(x = year, y = life_exp)) +
  geom_point()


# note: you don't need to use the $ sign when using ggplot
# geom_point is a scatter plot 
# ggplot is a function
# geom_histo, geom_bar, geom_boxplot, = are all the diff types of plots you can create with ggplot
```

The call to `ggplot` and `aes` sets up the basics of how we are going to represent the various columns of the data frame. `aes` defines the "aesthetics", which is how columns of the data frame map to graphical attributes such as x and y position, color, size, etc. `aes` is another example of magic "non-standard evaluation", arguments to `aes` may refer to columns of the data frame directly. We then literally add layers of graphics ("geoms") to this.

Further aesthetics can be used. Any aesthetic can be either numeric or categorical, an appropriate scale will be used.

```{r warning=F}
a <- ggplot(data = gap_geo, 
       aes(x=year, y=life_exp, color=region, size=population)) +
    geom_point()
a
```

### Challenge: make a ggplot {.challenge}

This R code will get the data from the year 2010:

```{r warning=F}
head(gap_geo, 5)
gap2010 <- filter(gap_geo, year == 2010)
head(gap2010, 5) 

# OR you can write the above code like this:
# gap_geo[gap_geo$year == 2010]
# gap2010
```

Create a ggplot of this with:

-   `gdp_percap` as x.
-   `life_exp` as y.
-   `population` as the size.
-   `region` as the color.

```{r warning=F}
# the line just above could be: ```{r warning=F, fig.width=2, fig.height=3} # to change the size of the figure
# the line just above could also be: ```{r warning=F, eval = False}

# Assign ggplot to variable for downstream challenges
ggplot(data = gap2010, 
       aes(x = gdp_percap, y = life_exp, size = population, color = region)) +
  geom_point()
```

## Further geoms

To draw lines, we need to use a "group" aesthetic.

```{r warning=F}
a <- ggplot(gap_geo, aes(x=year, y=life_exp, group=name, color=region)) +
    geom_line()
a
```

A wide variety of geoms are available. Here we show Tukey box-plots. Note again the use of the "group" aesthetic, without this ggplot will just show one big box-plot.

```{r warning=F}
ggplot(gap_geo, aes(x=year, y=life_exp, group=year)) +
    geom_boxplot()

# as.factor method
ggplot(gap_geo, aes(x=as.factor(year), y=life_exp)) +
    geom_boxplot()
```

`geom_smooth` can be used to show trends.

```{r warning=F}
ggplot(gap_geo, aes(x=year, y=life_exp)) +
    geom_point() +
    geom_smooth()
```

Aesthetics can be specified globally in `ggplot`, or as the first argument to individual geoms. Here, the "group" is applied only to draw the lines, and "color" is used to produce multiple trend lines:

```{r warning=F}
ggplot(gap_geo, aes(x=year, y=life_exp)) +
    geom_line(aes(group=name)) +
    geom_smooth(aes(color=oecd))
```

## Highlighting subsets

Geoms can be added that use a different data frame, using the `data=` argument.

```{r warning=F}
gap_australia <- filter(gap_geo, name == "Australia")
gap_vietnam <- filter(gap_geo, name == "Vietnam")

ggplot(gap_geo, aes(x = year, y = life_exp, group = name, color = region)) +
    geom_line() +
    geom_line(data = gap_australia, color="red", size=2) + # you can use source codes to make the plot lines colors of your own choice
    geom_line(data = gap_vietnam, color='blue', size =2) # note how you can't put gap_aus and gap_vietnam on the same geom_line 

# gghighlight from Slack
```

Notice also that the second `geom_line` has some further arguments controlling its appearance. These are **not** aesthetics, they are not a mapping of data to appearance, but rather a direct specification of the appearance. There isn't an associated scale as when color was an aesthetic.

## Fine-tuning a plot

Adding `labs` to a ggplot adjusts the labels given to the axes and legends. A plot title can also be specified.

```{r warning=F}
ggplot(gap_geo, aes(x=year, y=life_exp)) +
    geom_point() +
    labs(x="Year", y="Life expectancy (Age)", title="Gapminder") # have to make sure labels and titles are strings so R doesn't get confused and thing it is another variable. 
```

Now, the figure has proper labels and titles. However, the title is not at the center of the figure. We can further customize it using `theme()` function (for more detail please see the docs `?theme`).

```{r warning=F}
ggplot(gap_geo, aes(x=year, y=life_exp)) +
    geom_point() +
    labs(x="Year", y="Life expectancy (Age)", title="Gapminder") +
    theme(plot.title = element_text(hjust = 0.5),
          axis.text.x = element_blank(),
          panel.background = element_rect(fill = "white"))

# to move the title to the center => adjust it horizontally = using hjust 
# element_blank function removes all elements on the x-axis 
# element_rect changes the background colour of the plot  
# another way to do all these things is to use the diff theme overlays ava with ggplot (which automatically do the adjustments) eg: theme_dark, theme_classic ...etc More can also be downloaded online. 

```

`coord_cartesian` can be used to set the limits of the x and y axes. Suppose we want our y-axis to start at zero.

```{r warning=F}
ggplot(gap_geo, aes(x=year, y=life_exp)) +
    geom_point() +
    coord_cartesian(ylim=c(10,60))
```

Type `scale_` and press the tab key. You will see functions giving fine-grained controls over various scales (x, y, color, etc). These allow transformations (eg log10), and manually specified breaks (labelled values). Very fine grained control is possible over the appearance of ggplots, see the ggplot2 documentation for details and further examples.

### Challenge: refine your ggplot {.challenge}

Continuing with your scatter-plot of the 2010 data, add axis labels to your plot.

Give your x axis a log scale by adding `scale_x_log10()`.

```{r}
ggplot(data = gap2010, 
       aes(x = gdp_percap, y = life_exp, size = population, color = region)) +
  geom_point() +
  labs(title = "Linear")

ggplot(data = gap2010, 
       aes(x = gdp_percap, y = life_exp, size = population, color = region)) +
  geom_point() +
  scale_x_log10() + # this is to scale it by log
  labs(title = "Log") +
  annotation_logticks(side = 'b')
```

## Faceting

Faceting lets us quickly produce a collection of small plots. The plots all have the same scales and the eye can easily compare them.

```{r warning=F}
ggplot(gap_geo, aes(x=year, y=life_exp, group=name)) +
    geom_line() +
    facet_wrap(~region) +
  labs(title = "Regions")

ggplot(gap_geo, aes(x=year, y=life_exp, group=name)) +
    geom_line() +
    facet_wrap(~oecd) +
  labs(title = "OECD")
```

Note the use of `~`, which we've not seen before. `~` syntax is used in R to specify dependence on some set of variables, for example when specifying a linear model. Here the information in each plot is dependent on the continent.

### Challenge: facet your ggplot {.challenge}

Let's return again to your scatter-plot of the 2010 data.

Adjust your plot to now show data from all years, with each year shown in a separate facet, using `facet_wrap(~ year)`.

Advanced: Highlight Australia in your plot.

```{r, warning=F, fig.height=10, fig.width=10}
ggplot(data = gap_geo, 
       aes(x = gdp_percap, y = life_exp, size = population, color = region)) +
  geom_point(size = 1) +
  facet_wrap(~year)
```

# Dealing with **Missing Values** when plotting

```{r, eval=FALSE}
# run once to install package: 
install.packages("naniar")
```

```{r}
library(naniar)
```

In R, `NA` (or not available) is used to represent missing values which often interfere with downstream data analysis. Here we will look at a few ways to explore and handle missingness in our data. This example uses the `naniar` package, however base functions like `is.na()` can also be used.

```{r}
var1 <- c(1, 2, NA, 4:10, NA, 12:15)
is.na(var1)

# to remove NAs
var1[!is.na(var1)]
```

Now let us look at `airquality`, a built-in data set of daily air quality measurements in New York, from May to September 1973.

```{r}
head(airquality)
```

To check if `NA`s are present in a vector use the function `any_na()`

```{r}
any_na(airquality$Ozone)

# Challenge: In the airquality dataset, check if there are any NA's in the "Wind" column.
any_na(airquality$Wind)
```

The `gg_miss_var()` and `vis_miss()` functions can be used to provide a visual output

```{r}
gg_miss_var(airquality) # plots the amount of missing values in each column of our data frame
vis_miss(airquality) # creates a heat-map showing where the missing N/a values in our columns are. This is a useful function to figure our where your data is in bigger data frames. 
```

Calculate the mean and standard deviation of solar radiation - `Solar.R`. Does the output make sense?

```{r}
mean(airquality$Solar.R)
sd(airquality$Solar.R)

# the ans is an error because the N/a values don't make sense to R
# so we have to use the na.rm function to tell R to ignore the N.a values 
```

Now set the argument `na.rm` as TRUE

```{r}

mean(airquality$Solar.R, na.rm = TRUE)
sd(airquality$Solar.R, na.rm = TRUE)
```

To remove NA values use the base function `na.omit()`

```{r}
newdf <- na.omit(airquality)

# what do you observe?
vis_miss(newdf)
```

**NOTE** - be careful when using `na.omit()` as it removes all observations with `NA`s

Now lets see what using na.omit() did to the "Solar.R" values

```{r}
ggplot(newdf,
       aes(x = Month,
           y = Day,
           fill = Solar.R)) +
  geom_tile()
```

From the plot we can see because na.omit() removes all observations with NAs, many of the observations are removed even though they are present in the airquality dataset.

If we want to avoid losing additional data we can use ggplot to create a plot and just ignore the NA values by colouring them grey (or another colour) to retain all the available values in the dataset.

```{r}
ggplot(airquality,
       aes(x = Month,
           y = Day,
           fill = Solar.R)) +
  geom_tile() +
  scale_fill_continuous(na.value = NA) # to remove the NA tiles
# you can also replace Na with "red" to color the NA tiles red
```

## Saving ggplots

The act of plotting a ggplot is actually triggered when it is printed. In an interactive session we are automatically printing each value we calculate, but if you are using it with a programming construct such as a for loop or function you might need to explicitly `print( )` the plot.

GGplots can be saved using `ggsave`.

```{r, eval=FALSE}
# Plot created but not shown.
p <- ggplot(gap_geo, aes(x=year, y=life_exp)) + geom_point()
# Only when we try to look at the value p is it shown
p
# Alternatively, we can explicitly print it
print(p)
# To save to a file
ggsave(filename = "./Plot_Output/test.png", plot = p,
       width = 8,
       height = 8,
       dpi = 600,
       device = 'png')

# This is an alternative method that works with "base R" plots as well:
png("./Plot_Output/test.png")
print(p)
dev.off()
```

### Tip about sizing {.tip}

Figures in papers tend to be quite small. This means text must be proportionately larger than we usually show on screen. Dots should also be proportionately larger, and lines proportionately thicker. The way to achieve this using `ggsave` is to specify a small width and height, given in inches. To ensure the output also has good resolution, specify a high dots-per-inch, or use a vector-graphics format such as PDF or SVG.

```{r, eval=FALSE}
ggsave("test2.png", p, width=3, height=3, dpi=600)
```
